Quantum H theorem, variational entropy, and relaxation processes

### 2021-03-10 tex/relaxproc.tex
Cambridge corrections applied.

In order to obtain a time-evolution equation for an out-of-equilibrium quantum gas,
we propose the following approach. We start by evaluating $\Delta\mathcal{H}=
\mathcal{H}(t_2)-\mathcal{H}(t_1)$,
where our quantum $H$-functional ---Eq.~(\ref{eq:qHdef})--- is evaluated at diferent times $t_1$ and $t_2$, with
$t_2 > t_1$. This yields:
%
\begin{eqnarray}\label{eq:deltaHdef}
	\Delta\mathcal H &=& \sum_{M=1}^{K}\sum_n [f''_{nM} \ln f''_{nM} - f'_{nM}	\ln f'_{nM}\nonumber \\ 
			&&\pm (1\mp f''_{nM}) \ln(1\mp f''_{nM}) \mp (1\mp f'_{nM}) \ln (1\mp f'_{nM})]\delta V_M. 
\end{eqnarray}
%
In the above equation, and for the rest of this section, we use the short-hand notation
$f_{nM}(t_2)\equiv f''_{nM}$ and 
$f_{nM}(t_1) \equiv f'_{nM}$.
Subsequently, in Eq.~(\ref{eq:deltaHdef}), we replace the distribution functions by their expresions in terms
of deviations from equilibrium ---Eq.~(\ref{eq:qFirstOrd})---, which renders:
\begin{eqnarray}\label{eq:deltaH1}
	\Delta\mathcal H &=& \sum_{M=1}^{K} \sum_n\Big[\bar{f}_n(1+g''_{nM}) \ln \bar{f}_{n}(1+g''_{nM})  -  
			\bar{f}_n(1+g'_{nM}) \ln \bar{f}_n(1+g'_{nM})\Big. \nonumber \\
		    &&  \pm \big(1 \mp \bar{f}_n \{1+g''_{nM}\}\big) \ln \big(1\mp \bar{f}_n\{1+g''_{nM} \}\big)\nonumber \\
		    &&  \mp \big(1\mp \bar{f}_n \{ 1+g'_{nM} \}\big) \ln\big(1\mp \bar{f}_n \{1+g'_{nM} \}\big)\Big.\Big]
		    \delta V_M.
\end{eqnarray}
Afterwards, we expand the logarithmic terms up to the first-order in $g'_{nM}$ and $g''_{nM}$
and rearrange the result, which gives:
%
\begin{equation}\label{eq:deltaH2}
\Delta\mathcal H = \sum_{M}^{K}\sum_n [\bar{f}_n(1
	+\ln \bar{f}_n)-\bar{f}_n \{\ln(1\mp \bar{f}_n)\mp 1 \}](g''_{nM}-g'_{nM})\delta V_M.
\end{equation}
%
Finally we divide Eq.~(\ref{eq:deltaH2}) by $\Delta t\equiv t_2-t_1$, and take the limit
$\Delta t\to0$ to obtain:
%
\begin{equation}\label{eq:deltaH3}
\frac{d\mathcal H}{dt} = \sum_{M}^{K}\sum_n [\bar{f}_n(1
	+\ln \bar{f}_n)-\bar{f}_n \{\ln(1\mp \bar{f}_n)\mp 1 \}]\left(\frac{g_{nM}}{dt}\right)\delta V_M.
\end{equation}
%

Eq.~(\ref{eq:deltaH3}) is, within our framework, the time-evolution equation for $g_{nM}$.
Clearly, for describing a concrete situation, it is required to provide a specific approximation for the deviation function $g_{nM}$, however this will be the subject of future work.

  %----------------------------------------

%-------------------------------------------------

>>>>

To obtain a time-evolution equation for an out-of-equilibrium quantum gas,
we propose the following approach. We start by evaluating $\Delta\mathcal{H}=
\mathcal{H}(t_2)-\mathcal{H}(t_1)$,
where our quantum $H$-functional---Eq.~(\ref{eq:qHdef})---is evaluated at different times $t_1$ and $t_2$, with
$t_2 > t_1$. This yields
%
\begin{eqnarray}\label{eq:deltaHdef}
	\Delta\mathcal H &=& \sum_{M=1}^{K}\sum_n [f''_{nM} \ln f''_{nM} - f'_{nM}	\ln f'_{nM}\nonumber \\ 
			&&\pm (1\mp f''_{nM}) \ln(1\mp f''_{nM}) \mp (1\mp f'_{nM}) \ln (1\mp f'_{nM})]\delta V_M. 
\end{eqnarray}
%
In the above equation, and for the rest of this section, we use the short-hand notation
$f_{nM}(t_2)\equiv f''_{nM}$ and 
$f_{nM}(t_1) \equiv f'_{nM}$.
Subsequently, in Eq.~(\ref{eq:deltaHdef}), we replace the distribution functions with their expressions in terms
of deviations from equilibrium---Eq.~(\ref{eq:qFirstOrd})---which renders:
\begin{eqnarray}\label{eq:deltaH1}
	\Delta\mathcal H &=& \sum_{M=1}^{K} \sum_n\Big[\bar{f}_n(1+g''_{nM}) \ln \bar{f}_{n}(1+g''_{nM})  -  
			\bar{f}_n(1+g'_{nM}) \ln \bar{f}_n(1+g'_{nM})\Big. \nonumber \\
		    &&  \pm \big(1 \mp \bar{f}_n \{1+g''_{nM}\}\big) \ln \big(1\mp \bar{f}_n\{1+g''_{nM} \}\big)\nonumber \\
		    &&  \mp \big(1\mp \bar{f}_n \{ 1+g'_{nM} \}\big) \ln\big(1\mp \bar{f}_n \{1+g'_{nM} \}\big)\Big.\Big]
		    \delta V_M.
\end{eqnarray}
Subsequently, we expand the logarithmic terms up to the first-order in $g'_{nM}$ and $g''_{nM}$
and rearrange the result, which gives
%
\begin{equation}\label{eq:deltaH2}
\Delta\mathcal H = \sum_{M}^{K}\sum_n [\bar{f}_n(1
	+\ln \bar{f}_n)-\bar{f}_n \{\ln(1\mp \bar{f}_n)\mp 1 \}](g''_{nM}-g'_{nM})\delta V_M.
\end{equation}
%
Finally we divide Eq.~(\ref{eq:deltaH2}) by $\Delta t\equiv t_2-t_1$, and take the limit
$\Delta t\to0$ to obtain
%
\begin{equation}\label{eq:deltaH3}
\frac{d\mathcal H}{dt} = \sum_{M}^{K}\sum_n [\bar{f}_n(1
	+\ln \bar{f}_n)-\bar{f}_n \{\ln(1\mp \bar{f}_n)\mp 1 \}]\left(\frac{g_{nM}}{dt}\right)\delta V_M.
\end{equation}
%

Eq.~(\ref{eq:deltaH3}) is, within our framework, the time-evolution equation for $g_{nM}$.
Clearly, to describe a realistic situation, providing a specific approximation for the deviation function $g_{nM}$ is required. This subject will be explored in future work.
